---
title: "Assessment"
output: html_document
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exploring the pattern of prescription  of MAOIs (Monoamine-oxidase inhibitors)  

Monoamine-oxidase inhibitors are a class of antidepressant medications that is now rarely used. This is largely due to their ability to interact with a lot of other medications as well as various foods. This means they are largely reserved for specialist prescriptions and are only used for treatment-resistant depression.

I intend to explore the prescription patterns based on geographical factors, age group (potentially), sex and deprivation. I also want to compare data for a couple of months to see if there is any obvious difference in patterns (although, I don't expect any).

"warning=FALSE, message=FALSE" attribute is used in order not to print unnecessary console output
## This section loads the necessary libraries
```{r warning=FALSE, message=FALSE }
library(tidyverse) 
library(janitor) # cleaning data
library(gt) # tables
library(here) # directory structure 
```
## This section loads the necessary datasets.

Most of them are loaded as is for now in case I need more data. Once I am fully dome, I will be able to load only selected columns that I use to reduce the "hoarding"
``` {r warning=FALSE, message=FALSE}
#Loading Healthboard names and codes only as a reference dataset 

HB_lookup <- read_csv("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv") %>% 
  clean_names() %>% 
  select(hb, hb_name)

#Loading GP practices details (including addresses)  

gp_addresses <- read_csv("https://www.opendata.nhs.scot/dataset/f23655c3-6e23-4103-a511-a80d998adb90/resource/30b06220-17ad-44e8-b6c5-658d41ec1ea5/download/practice_contactdetails_jul2025-open-data.csv") %>%
  clean_names()

#Loading GP practices population details (including age and sex distribution)

gp_practice_populations <- read_csv("https://www.opendata.nhs.scot/dataset/e3300e98-cdd2-4f4e-a24e-06ee14fcc66c/resource/9cb891c1-e886-4470-b06c-9a8cf6bf4b22/download/practice_listsizes_jul2025-open-data.csv") %>% 
  clean_names()

#Loading Scottish index of multiple deprivation data

simd <- read_csv("https://www.opendata.nhs.scot/dataset/78d41fa9-1a62-4f7b-9edb-3e8522a93378/resource/acade396-8430-4b34-895a-b3e757fa346e/download/simd2020v2_22062020.csv") %>% 
  clean_names()
```
## This section deals with the initial data exploration from June 2025 and rendering the first graph
```{r warning=FALSE, message=FALSE}
#Loading June 2025 GP prescriptions data

data_jun_2025 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/0f2931af-3d08-4a6b-92cf-b61be4ce8d97/download/pitc202506.csv") %>% 
  clean_names()

#Creating a vector of the 4 MAOIs used in the UK

maois <- c("TRANYLCYPROMINE", "PHENELZINE", "ISOCARBOXAZID", "MOCLOBEMIDE")


#Sorting out the records relevant to MAOIs into a separate dataset (done this way in case I want to use something from the main prescriptions dataset later; will likely combine with the line above when finished - general pattern for most code here)

maoi_data <- data_jun_2025 %>% 
  filter(str_detect(bnf_item_description, paste(maois, collapse = "|")))

#Joining the healthboard names to the MAOIs dataset and filtering out the columns of interest

maoi_data <- maoi_data %>% 
  left_join(HB_lookup, join_by(hbt == hb))%>% 
  select(hb_name, hbt, gp_practice, bnf_item_description, paid_quantity, paid_date_month)

#Collapsing the data so that all MAOIs are combines (worth noting that PHENELZINE isn't present at all)
maoi_data_summarised <- maoi_data %>% 
  group_by(hb_name, hbt, gp_practice, paid_date_month) %>% 
  summarise(drug_quantity = sum(paid_quantity))

#Adding the GP practice addresses

maoi_data_summarised <- maoi_data_summarised %>% 
  left_join(gp_addresses, join_by(gp_practice == practice_code)) %>% 
  select(hb_name:gp_practice_name, postcode, data_zone)

#Adding the population data

maoi_data_summarised <- maoi_data_summarised %>% 
  left_join(gp_practice_populations, join_by(gp_practice == practice_code)) %>% 
  select(hb_name:data_zone, sex, all_ages)

#Creating a bar chart of the association between healthboard and the number of MAOI prescriptions per person grouped by sex

maoi_data_summarised %>% 
  filter(!is.na(all_ages)) %>% 
  ggplot(aes(x = hb_name, y = drug_quantity / all_ages, fill = sex)) +
  geom_col() +
  facet_wrap(~sex) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Prescription of MAOIs per person in each health board grouped by sex",
       x = "Healthboard name",
       y = "MOAI quantity per person")

```
